---
title: "The toy example"
output: 
    html_notebook
---

# Packages: make sure they are installed

```{r pack}
library(tidyverse)
library(rpart)
library(rpart.plot)
library(gridExtra)
```


# Toy example: setup

We run simulations and create the first graph.

```{r lin_quad, message = FALSE, warning = FALSE}
sample <- 10^3                                                       # Sample size
x1 <- runif(sample)                                                  # x1 variable
x2 <- runif(sample)                                                  # x2 variable
y <- x1/3 + sin(10*x2)/4 + x2/9 - 0.12 + rnorm(sample)/10
data <- data.frame(y, x1, x2)                                        # Aggregate dataset
data %>% ggplot() +                                                  # Plot
    geom_point(aes(x = x1, y = y), color = "blue2", size = 0.3, alpha = 0.4) +
    geom_point(aes(x = x2, y = y), color = "red2", size = 0.3, alpha = 0.4) +
    stat_smooth(aes(x = x1, y = y), color = "blue2") +
    stat_smooth(aes(x = x2, y = y), color = "red2") +
    xlab("x1 (blue) or x2 (red)")
```


# Gain analysis (pre-filter)

We compute the important indicators stemming from splits (the gains essentially).

```{r gains, message = FALSE, warning = FALSE}
sample <- 10^6                                                       # More points & then start again.
x1 <- runif(sample)                                                  # x1 variable
x2 <- runif(sample)                                                  # x2 variable
y <- x1/3 + sin(10*x2)/4 + x2/9 - 0.12 + rnorm(sample)/10
data <- data.frame(y, x1, x2) 

gain <- function(fit){     # Simple function to extract the gains
    out_fit <- fit$frame
    var10 <- out_fit$dev[1]
    var11 <- out_fit$dev[2]  
    var12 <- out_fit$dev[3]
    return((var10-var11-var12)/out_fit$n[1])
}

fit1 <- rpart(y ~ x1, data = data, maxdepth = 1)
gain(fit1) # Gain realized by first variable
fit2 <- rpart(y ~ x2, data = data, maxdepth = 1)
gain(fit2) # Gain realized by second variable
```

The second variable leads to the highest gain: this is the one that would be chosen.

And plot the corresponding trees.
```{r tree_prefilter}
layout(matrix(c(1,2), 1, 2, byrow = TRUE))
rpart.plot(fit1)
rpart.plot(fit2)
```

# Tree post-filter

We show what happens if you filter the data upfront: the split changes.

```{r tree_filter}
q <- 0.15
sample <- 10^4                                                       # Generate again.
x1 <- runif(sample)                                                  # x1 variable
x2 <- runif(sample)                                                  # x2 variable
y <- x1/3 + sin(10*x2)/4 + x2/9 - 0.12 + rnorm(sample)/10
data <- data.frame(y, x1, x2) 
data_filter <- data %>% filter(y < quantile(y,q) | y > quantile(y, 1-q))
fit <- rpart(y ~ x1 + x2, data = data_filter, maxdepth = 1)  
rpart.plot(fit)
```

# Graphical explanation

This one is a bit long, we keep a lot of information. We show the separations dictated by the splits via a vertical line.

```{r filter, message = FALSE, warning = FALSE}
fit1 <- rpart(y ~ x1, data = data_filter, maxdepth = 1)
fit2 <- rpart(y ~ x2, data = data_filter, maxdepth = 1)
split1 <- fit1$splits[fit1$splits[,3]==min(fit1$splits[,3]),4]
split2 <- fit2$splits[fit2$splits[,3]==min(fit2$splits[,3]),4]
data_filter <- data %>% filter(y < quantile(y,q) | y > quantile(y, 1-q))

g1 <- data_filter %>% 
    ggplot() +
    geom_point(aes(x = x1, y = y), color = "blue", size = 0.3) +
    stat_smooth(aes(x = x1, y = y)) +
    annotate("segment", x = split1, xend = 1, 
             y = mean(data_filter$y[data_filter$x1>split1]), 
             yend = mean(data_filter$y[data_filter$x1>split1]), 
             colour = "black", linetype = 2, size = 1.5) +
    annotate("segment", x = 0, xend = split1, 
             y = mean(data_filter$y[data_filter$x1<split1]), 
             yend = mean(data_filter$y[data_filter$x1<split1]), 
             colour = "black", linetype = 2, size = 1.5) +
    annotate("text", x = split1-0.05, y = 0.6 , label = "split", color = "red") +
    geom_vline(xintercept = split1, color = "red") +
    ggtitle("First variable")

g2 <- data_filter %>% 
    ggplot() +
    geom_point(aes(x = x2, y = y), color = "red", size = 0.3) +
    stat_smooth(aes(x = x2, y = y)) +
    annotate("segment", x = split2, xend = 1, 
             y = mean(data_filter$y[data_filter$x2>split2]), 
             yend = mean(data_filter$y[data_filter$x2>split2]), 
             colour = "black", linetype = 2, size = 1.5) +
    annotate("segment", x = 0, xend = split2, 
             y = mean(data_filter$y[data_filter$x2<split2]), 
             yend = mean(data_filter$y[data_filter$x2<split2]), 
             colour = "black", linetype = 2, size = 1.5) +
    annotate("text", x = split2-0.05, y = 0.6 , label = "split", color = "red") +
    geom_vline(xintercept = split2, color = "red") +
    ggtitle("Second variable")
grid.arrange(g1,g2)
```