---
title: "ToT Formal plots"
output: html_notebook
---


# Gains: linear versus quadratic generators

First, we defined a few functions we use to compare gains in simple cases. They are dedicated to the evaluation of gain in simple parametric models: linear and quadratic generators for the conditional average. At the end of the chunck, we compare the relative gains from one generator to another under an iso-variance constraint.

```{r functions}
c_plus <- function(b){                          # Split point for quadratic generator (1)
    return((1+6*b+sqrt(17-36*b+36*b^2))/8)
}
c_minus <- function(b){                         # Split point for quadratic generator (2)
    return((1+6*b-sqrt(17-36*b+36*b^2))/8)
}
init_var_lin <- function(a){                    # Total variance from linear generator
    return(a^2/12)
}
init_var_quad <- function(b){                   # Total variance from quadratic generator
    return((15*b^2-15*b+4)/45)
}
V_quad <- function(c, b){                       # Post-split variance for the quadratic generator
    return(  (4 - 15*b + 15*b^2 - 5*c + 30*b*c - 45*b^2*c - 5*c^2 + 45*b^2*c^2 + 5*c^3 - 30*b*c^3 + 5*c^4)/45)
}
gain_quad <- function(b){                       # Gain obtained by the quadratic generator
    return(init_var_quad(b) - V_quad(c_plus(b),b))
}
gain_lin <- function(b){                        # Gain obtained by the linear generator under iso-variance
    return((15*b^2-15*b+4)/60)
}
gain_diff <- function(b){                       # Difference between gains
    gain_lin(b)- gain_quad(b)
}

data.frame(x = c(0,0.49)) %>%                   # Plot
    ggplot(aes(x = x)) + 
    stat_function(fun = gain_diff) +
    geom_segment(aes(x = 0.383, y = -0.0024, xend = 0.383, yend = 0), linetype = 2) +
    ylim(-0.0024,0.00215)
```


# Plots of parametric generators

Below, we define the families of generators that we work with in the simulations (the piecewise constant is nonetheless omitted in the paper).

```{r graphs}
generator <- function(x, type, c){
    if(type == "linear"){return(c*x)}
    if(type == "quad"){return((x-c)^2)}
    if(type == "exp"){return(exp(-x*c))}
    if(type == "log"){return(log(x+c))}
    if(type == "sin"){return(sin(x*c)/c)}
    if(type == "outlier"){return(c*(x-0.5)^3)}
    if(type == "piece"){                            # Piecewise constant
        weightz <- arima.sim(list(ar = c(0.99)), c)
        div <- length(weightz)
        weightz <- (weightz - mean(weightz)) / sd(weightz) / runif(1, min = 8, max = 13)
        if(length(x) == 1){                         # Real input (dim = 1)
            inter <- seq(0, 1-1/div, by = 1/div)    # Intervals
            ind <-  sum(x > inter)
            return(weightz[ind] - mean(weightz))
        } else {                                    # Vector input
            inter <- rep(seq(0, 1-1/div, by = 1/div), length(x)) %>%
                matrix(ncol = div, byrow = T)
            comp <- rep(x, div) %>%
                matrix(ncol = div, byrow = F)
            ind <- rowSums(x > inter)
            return(weightz[ind])
        }
    }
    if(type == "poly"){                              # Polynomial generator
        pars <- rnorm(c) 
        pars <- pars - mean(pars) # To make sure that the behavior is not too montonous
        m <- sum(pars/(2:(c+1)))
        x <- sapply(x, `^`, 1:c)
        p <- matrix(rep(pars, ncol(x)), nrow = c, byrow = F)
        out <- colSums(p*x) - m
        return(out / sd(out) / runif(1, min = 12, max = 21))
    }
}

m_gen <- function(type, c){                           # Average values
    if(type == "linear") return(c/2) 
    if(type == "quad") return(1/3-c+c^2)
    if(type == "exp") return((1-exp(-c))/c)
    if(type == "log") return(-1 - c*log(c)+(1+c)*log(1+c))
    if(type == "sin") return((1-cos(c))/c^2)
    if(type == "outlier"){return(0)}                  # A true zero
    if(type == "piece") {return(0)}                   # Not defined (parameter dependent)
    if(type == "poly") {return(0)}                    # Not defined (parameter dependent)
}

fin_gen2 <- function(x, type,cc){                     # Final generator: with mean retrieved
    return(generator(x, type, cc) - m_gen(type, cc))
}

ggplot(data.frame(x = c(0.01,1)), aes(x = x)) +
    stat_function(fun = fin_gen2, args = list(type = "linear", c = 0.1), aes(color = "linear")) +
    stat_function(fun = fin_gen2, args = list(type = "linear", c = 0.2), aes(color = "linear")) +
    stat_function(fun = fin_gen2, args = list(type = "linear", c = 0.3), aes(color = "linear")) +
    stat_function(fun = fin_gen2, args = list(type = "quad", c = 0.51), aes(color = "quad")) +
    stat_function(fun = fin_gen2, args = list(type = "quad", c = 0.55), aes(color = "quad")) +
    stat_function(fun = fin_gen2, args = list(type = "quad", c = 0.59), aes(color = "quad")) +
    stat_function(fun = fin_gen2, args = list(type = "exp", c = 0.1), aes(color = "exp")) +
    stat_function(fun = fin_gen2, args = list(type = "exp", c = 0.2), aes(color = "exp")) +
    stat_function(fun = fin_gen2, args = list(type = "exp", c = 0.3), aes(color = "exp")) +
    stat_function(fun = fin_gen2, args = list(type = "log", c = 3), aes(color = "log")) +
    stat_function(fun = fin_gen2, args = list(type = "log", c = 8), aes(color = "log")) +
    stat_function(fun = fin_gen2, args = list(type = "log", c = 13), aes(color = "log")) +
    stat_function(fun = fin_gen2, args = list(type = "sin", c = 7), aes(color = "sin")) +
    stat_function(fun = fin_gen2, args = list(type = "sin", c = 10), aes(color = "sin")) +
    stat_function(fun = fin_gen2, args = list(type = "sin", c = 14), aes(color = "sin")) +
    stat_function(fun = fin_gen2, args = list(type = "poly", c = 6), aes(color = "poly")) +
    stat_function(fun = fin_gen2, args = list(type = "poly", c = 6), aes(color = "poly")) +
    stat_function(fun = fin_gen2, args = list(type = "poly", c = 6), aes(color = "poly")) 
#ggsave("generators.pdf")
```


# Counterexamples

Finally, some counterexamples. One is ok, the second is pathological.

```{r counter}
library(gridExtra)
g1 <- ggplot(data.frame(x = c(0,1)), aes(x = x)) + stat_function(fun = function(x) (x-0.5)^3) 
g2 <- ggplot(data.frame(x=1:8, y = c(1,1,2,6,-5,-1,-2,-2))) + geom_bar(aes(x = x, y = y/10), stat = "identity")
grid.arrange(g1, g2, ncol=2) #%>% ggsave(filename = "counter.pdf")
```

